A Dissection of Foregrounds in Epoch of Reionization Observations from the Murchison Widefield Array

### 2015-01-17 ms.tex
New section on role of antenna shapes partially complete

{\bf Interestingly, \citet{thy13} and \citet{pob13} have found a similar feature in their simulations and data respectively. In the latter, this feature is described as ``edge brightening'' and attributed to the Galactic plane near the horizon. From their discussion, it is unclear what fraction of power in that feature arises from ubiquitous diffuse emission unlike in our {\it zenith} pointing where the Galactic plane is not expected to play a significant role. On the other hand, \citet{thy13} statistically modeled point sources below the confusion limit and attributed the feature to an increase in solid angles subtended by delay bins closer to the horizon limits thereby, enhancing the resulting integrated emission. This must be true for diffuse emission as well. Thus we argue the feature arises primarily due to an increase in sensitivity of long baselines to diffuse emission from far off--axis angles caused by foreshortening, and partly due to a higher integrated emission in delay bins close to the horizon limits which map larger solid angles in these directions}.

@@

This screening technique can be generalized to optimize between foreground mitigation and loss of sensitivity from discarding data. Figure~\ref{fig:screening} shows how the typical foreground contamination\footnote{Foreground contamination is measured by standard deviation of noiseless $V_{b\tau}$ from foregrounds in the {\it EoR window}.} in the {\it EoR window} depends on the orientations and lengths of discarded antenna spacings. We choose antenna spacings oriented eastward to varying degrees of directedness, i.e., $-7.5\arcdeg~\le\theta_\textrm{b} < 7.5\arcdeg$ (solid circles), $-15\arcdeg\le\theta_\textrm{b}< 15\arcdeg$ (solid squares), and $-22.5\arcdeg\le\theta_\textrm{b}< 22.5\arcdeg$ (solid stars). Among antenna spacings that satisfy these criteria, we discard data from those whose lengths are shorter than $|\boldsymbol{b}|_\textrm{max}$ ($x$--axis) and show foreground contamination estimated in the {\it EoR window} from all remaining antenna spacings. In other words, these plots demonstrate the progress in mitigation as orientation and maximum length of discarded antenna spacings are varied. % Foreground contamination estimated from only northward (dashed line) and eastward (dot--dashed line) antenna spacings represent the lower and upper limits, respectively, and are shown for reference. 
The fraction of discarded antenna spacings discarded relative to the total number is shown in dotted lines for different ranges of $\theta_\textrm{b}$. It is seen that foreground contamination can be mitigated by a factor between $\sim 1.5$ ($|\theta_\textrm{b}|\le$~7.5\arcdeg) and $\sim 10$ ($|\theta_\textrm{b}|\le$~22.5\arcdeg). The latter limit is achieved with a mere 5\% loss of data for $|\boldsymbol{b}|_\textrm{max}\simeq30$~m. And discarding antenna spacings with lengths $|\boldsymbol{b}|\gtrsim 30$~m does not mitigate foreground contamination any further and would only lead to loss of sensitivity as the fraction of discarded baselines increases from $\sim 5$\% to $\sim 25$\%. On the other hand, a bright point source at the same location will give rise to foreground contamination even on longer antenna spacings. Such cases will necessitate discarding more or, in the worst case, all of the eastward oriented antenna spacings. {\bf These results are specific to the MWA data. In general, the array configuration will be a significant factor in determining such thresholds}.

@@

>>>>

{\bf Interestingly, \citet{thy13} and \citet{pob13} have found a similar feature in their simulations and data respectively. In the latter, this feature is described as ``edge brightening'' and attributed to the Galactic plane near the horizon. From their discussion, it is unclear what fraction of power in that feature arises from ubiquitous diffuse emission unlike in our {\it zenith} pointing where the Galactic plane is not expected to play a significant role. On the other hand, \citet{thy13} statistically modeled point sources below the confusion limit and attributed the feature to an increase in solid angles subtended by delay bins closer to the horizon limits thereby, enhancing the resulting integrated emission. This must be true for diffuse emission as well. Thus we argue the feature arises primarily due to an increase in sensitivity of long baselines to diffuse emission caused by foreshortening towards angles far off--axis, where the increased size scales (or solid angles) containing higher levels of integrated emission and spatial scales given by the inverse of foreshortened baseline vectors become matched}.

@@

This screening technique can be generalized to optimize between foreground mitigation and loss of sensitivity from discarding data. Figure~\ref{fig:screening} shows how the typical foreground contamination\footnote{Foreground contamination is measured by standard deviation of noiseless $P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel)$ from foregrounds in the {\it EoR window}.} in the {\it EoR window} depends on the orientations and lengths of discarded antenna spacings. We choose antenna spacings oriented eastward to varying degrees of directedness, i.e., $-7.5\arcdeg~\le\theta_\textrm{b} < 7.5\arcdeg$ (solid circles), $-15\arcdeg\le\theta_\textrm{b}< 15\arcdeg$ (solid squares), and $-22.5\arcdeg\le\theta_\textrm{b}< 22.5\arcdeg$ (solid stars). Among antenna spacings that satisfy these criteria, we discard data from those whose lengths are shorter than $|\boldsymbol{b}|_\textrm{max}$ ($x$--axis) and show foreground contamination estimated in the {\it EoR window} from all remaining antenna spacings. In other words, these plots demonstrate the progress in mitigation as orientation and maximum length of discarded antenna spacings are varied. % Foreground contamination estimated from only northward (dashed line) and eastward (dot--dashed line) antenna spacings represent the lower and upper limits, respectively, and are shown for reference. 
The fraction of discarded antenna spacings discarded relative to the total number is shown in dotted lines for different ranges of $\theta_\textrm{b}$. It is seen that foreground contamination can be mitigated by a factor between $\sim 1.5$ ($|\theta_\textrm{b}|\le$~7.5\arcdeg) and $\sim 10$ ($|\theta_\textrm{b}|\le$~22.5\arcdeg). The latter limit is achieved with a mere 5\% loss of data for $|\boldsymbol{b}|_\textrm{max}\simeq30$~m. And discarding antenna spacings with lengths $|\boldsymbol{b}|\gtrsim 30$~m does not mitigate foreground contamination any further and would only lead to loss of sensitivity as the fraction of discarded baselines increases from $\sim 5$\% to $\sim 25$\%. On the other hand, a bright point source at the same location will give rise to foreground contamination even on longer antenna spacings. Such cases will necessitate discarding more or, in the worst case, all of the eastward oriented antenna spacings. {\bf These results are specific to the MWA data. In general, the array configuration will be a significant factor in determining such thresholds}.

@@

\section{Role of Antenna Geometry}\label{sec:antenna-shape}

Having examined the role the foregrounds play in determining the delay spectrum obtained with an interferometer array, we turn our attention to another crucial factor, namely, the power pattern of the antenna (refer to equations~\ref{eqn:obsvis} and \ref{eqn:delay-transform}). In the following discussion, we assume there are no random delay fluctuations in the MWA tile beamforming process. The model power pattern so obtained is different from the one we used earlier, but for the present discussion it still serves as a good reference (Figure~\ref{fig:power_pattern}). Since the previous model suffices, we do not plot the new power pattern model obtained with ideal beamforming. We also assume an ideal flat bandpass shape over the 30.72~MHz bandwidth unlike the flagging of the edges of the coarse sub--bands described earlier in this paper, while still applying a {\it Blackman--Harris} window function.

\subsection{Uniformly Illuminated Sky}\label{sec:USM}

The brightness distribution on the sky was fixed to be a constant equal to the mean over the whole sky at 185~MHz with no variation along frequency. This helps isolate the signatures of the antenna power pattern from those of the foregrounds. Figure~\ref{fig:noiseless-usm-delay-spectrum} shows the delay spectrum power of such a sky model with the model power pattern of an MWA tile.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{fig14.eps}
\caption{Similar to Figure~\ref{fig:noiseless-dsm-delay-spectrum} but for a uniform sky brightness model. The features bear a close resemblance to those obtained with the diffuse emission model. The edge--heavy {\it two--pronged} features arise from foreshortening of baselines which increases their sensitivity to emission from larger size scales from near the horizon. The symmetries around $\tau=0$ or lack thereof correspond to symmetries in the power patterns of the respective pointings (see Figure~\ref{fig:power_pattern}). \label{fig:noiseless-usm-delay-spectrum}}
\end{figure}

Although not identical, these delay spectra power bear a close resemblance to those obtained with the diffuse sky model (see Figure~\ref{fig:noiseless-dsm-delay-spectrum}). This indicates that the delay spectra power seen from diffuse foregrounds is primarily explained by the delay spectrum signatures of the underlying power pattern, with the additional variations coming from anisotropic deviations of the diffuse foreground model from a uniformly illuminated sky.

The edge--heavy {\it two--pronged} features are evident, which are very similar to those seen earlier in the case of diffuse emission. As reasoned in \S\ref{sec:diffuse-long-baselines}, the foreshortening of long baselines causes them to be sensitive to emission from near the horizon.

Branches of foreground power seen at positive delays and close to the negative horizon delay limit in the {\it off--zenith} pointing correspond to radiation received from the direction of the primary lobe and the far eastward sidelobe, respectively, in the power pattern. On the other hand, the symmetry of the delay spectrum around $\tau=0$ seen in the {\it zenith} pointing is as expected from a symmetric power pattern.

\subsection{Experiments with Antenna Geometry}\label{sec:antgeom}

\begin{figure*}[htb]
\centering
\subfloat[][Power patterns]{\label{fig:power-patterns}\includegraphics[width=\linewidth]{fig15a.eps}} \\
\subfloat[][Delay spectra power]{\label{fig:antennas-delay-spectra}\includegraphics[width=\linewidth]{fig15b.eps}}
\caption{}
\label{fig:multi-antenna}
\end{figure*}
